---
title: "Safe Mutation Plans"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Safe Mutation Plans}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
```

Mutations change data in Stash. These helpers separate preparing a request
from sending it:

1. Build a plan without contacting Stash.
2. Inspect the inputs and omitted fields.
3. Execute explicitly with `dry_run = FALSE`.
4. Keep source indexes and failures with the results.

The helpers do not choose metadata, match objects, retry requests, or provide
transactions. The builder remains responsible for those decisions.

## The workflow

```r
prepare_mutations(data, build_input, na = "omit", null = "omit", operation = NULL)
execute_mutations(plan, mutate, dry_run = TRUE, on_error = "stop", progress = TRUE)
```

`prepare_mutations()` calls `build_input(row, index)` once per source row. The
builder must return the complete named GraphQL input, for example
`list(id = row$id, title = row$title)` for `galleryUpdate()`.

`execute_mutations()` calls the supplied mutation function as
`mutate(input = input)`. The default is a dry run, so the mutation function is
not called until `dry_run = FALSE` is supplied.

## A dry run

This example uses fake data and a fake mutation function, so it does not need a
Stash connection. The source uses `id`, the normal identifier column returned
by generated `find*()` functions.

```{r dry-run}
library(stashapi)

updates <- list(
  list(id = "101", title = "First gallery"),
  list(id = "102", title = "Second gallery")
)

plan <- prepare_mutations(
  updates,
  build_input = function(row, index) {
    list(id = row$id, title = row$title)
  },
  operation = "galleryUpdate"
)

plan$entries[[1]]$input
plan$entries[[1]]$target
```

Previewing never calls the mutation function:

```{r preview}
preview <- execute_mutations(
  plan,
  mutate = function(input) stop("this must not run during a dry run"),
  progress = FALSE
)

preview$dry_run
vapply(preview$results, function(entry) entry$status, character(1))
```

## Building from query results

The source can be a data frame made from a Stash query. The builder chooses
which columns to send; other columns remain local-only.

```{r query-result-shape}
galleries <- data.frame(
  id = c("201", "202"),
  title = c("Example gallery A", "Example gallery B"),
  date = c("2026-01-01", NA_character_),
  stringsAsFactors = FALSE
)

plan <- prepare_mutations(
  galleries,
  build_input = function(row, index) {
    list(id = row$id, title = row$title, date = row$date)
  },
  operation = "galleryUpdate"
)

plan$entries[[2]]$input
plan$entries[[2]]$omitted
```

Generated `find*()` results use `id` for the object identifier. Bulk inputs
use `ids` because they address several objects in one request.

## Missing values

By default, scalar `NA` and `NULL` fields are omitted. Missing elements in a
vector are removed. Empty vectors are preserved because they may have API
meaning.

```{r missing-values}
plan <- prepare_mutations(
  list(list(id = "301")),
  build_input = function(row, index) {
    list(
      id = row$id,
      title = NA_character_,
      details = NULL,
      tag_ids = list(ids = c("41", NA_character_), mode = "ADD")
    )
  }
)

plan$entries[[1]]$input
plan$entries[[1]]$omitted
```

Use `na = "error"` or `null = "error"` when missing data should stop plan
construction:

```{r strict-values, error=TRUE}
prepare_mutations(
  list(list(id = "302")),
  function(row, index) list(id = row$id, title = NA_character_),
  na = "error"
)
```

`NULL` does not mean “clear this field”. Explicit clearing requires separate
GraphQL-null semantics and is intentionally not provided here.

## Nested update modes

Bulk mutation fields such as `performer_ids` and `urls` are nested inputs. The
builder supplies both the values and the operation mode:

```r
performer_ids = list(
  ids = row$performer_ids,
  mode = "ADD"
)
```

The helpers preserve this structure. They do not flatten it, change `ADD` to
`SET`, or infer a mode.

```{r nested-modes}
source <- list(
  list(
    ids = c("401", "402"),
    performer_ids = c("51", NA_character_),
    urls = "https://example.test/gallery/one"
  )
)

plan <- prepare_mutations(
  source,
  build_input = function(row, index) {
    list(
      ids = row$ids,
      organized = TRUE,
      performer_ids = list(ids = row$performer_ids, mode = "ADD"),
      urls = list(values = row$urls, mode = "SET")
    )
  },
  operation = "bulkImageUpdate"
)

plan$entries[[1]]$input$performer_ids
plan$entries[[1]]$input$urls
```

## Executing and handling failures

For a live request, pass the generated operation as `mutate`:

```{r live-execution, eval=FALSE}
result <- execute_mutations(
  plan,
  mutate = stashapi::galleryUpdate,
  dry_run = FALSE,
  on_error = "continue"
)
```

The following offline example demonstrates indexed failures:

```{r failures}
mutate <- function(input) {
  if (identical(input$id, "102")) stop("example validation failure")
  list(id = input$id, updated = TRUE)
}

result <- execute_mutations(
  plan = prepare_mutations(
    list(list(id = "101"), list(id = "102"), list(id = "103")),
    function(row, index) list(id = row$id)
  ),
  mutate = mutate,
  dry_run = FALSE,
  on_error = "continue",
  progress = FALSE
)

vapply(result$results, function(entry) entry$index, integer(1))
vapply(result$results, function(entry) entry$status, character(1))
result$results[[2]]$error$message
```

With the default `on_error = "stop"`, execution stops at the first failure.
The `stashapi_mutation_error` condition includes the failed index, original
error, and partial results. There is still no rollback.

## Bulk updates: one request per source row

`prepare_mutations()` can prepare existing `bulk*Update()` calls. The plan
entry represents one request, not one object inside that request.

For the gallery-to-image workflow, the user script first obtains the image IDs
belonging to each gallery. It can then create one source row per gallery:

```{r bulk-plan}
bulk_source <- list(
  list(
    ids = c("601", "602"),
    studio_id = "7",
    performer_ids = c("51", "52"),
    urls = "https://example.test/gallery/one",
    date = "2026-01-01"
  ),
  list(
    ids = "603",
    studio_id = "7",
    performer_ids = character(),
    urls = "https://example.test/gallery/two",
    date = NA_character_
  )
)

bulk_plan <- prepare_mutations(
  bulk_source,
  build_input = function(row, index) {
    list(
      ids = row$ids,
      organized = TRUE,
      urls = list(values = row$urls, mode = "SET"),
      studio_id = row$studio_id,
      performer_ids = list(ids = row$performer_ids, mode = "ADD"),
      date = row$date
    )
  },
  operation = "bulkImageUpdate"
)

bulk_plan$entries[[1]]$input
```

Preview, then execute after review:

```{r bulk-execution, eval=FALSE}
execute_mutations(bulk_plan, stashapi::bulkImageUpdate, progress = FALSE)

bulk_result <- execute_mutations(
  bulk_plan,
  stashapi::bulkImageUpdate,
  dry_run = FALSE,
  on_error = "continue"
)
```

Each result represents one `bulkImageUpdate()` request. A successful request
does not create independent success records for every image because the
GraphQL operation returns one request-level response.

## What remains in the user script

The builder does not decide:

- which gallery metadata to copy;
- how galleries and images are matched;
- whether a relationship uses `ADD` or `SET`;
- how paths and filenames are parsed;
- whether a failed mutation should be retried.

That is intentional. The helpers make the mutation boundary visible and
reviewable without turning `stashapi` into a studio-specific automation layer.
